\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing income inequality with tax data - Switzerland from 1917 to 2010}

### 2014-09-17 paper/subdocuments/results.Rnw
export reldist data to stata creating stata plot (reldist) minor edits to the text in results.Rnw

In this section we will present results following the strategy introducded in the last section. As describded this includes changeing possible concepts within one field (incomde defintion, statistical unit, measurement and population coverage) while holding other conceptional differences constant. To get the best possible comparison we rely on different data sources (FTA-tax tax data and HBS-survey data) and different time periodes. The presentation of results therefore includes a despriction of the uses data.

@@

We start with the effect of income definition on the assessment of income inequality. As described in the previous section income definitions are either closer to the market outcome or the the income, which shapes the possibility to consume (disposable income). With FTA-Tax data in-between income measures can be defined. This includes

@@

As \ref{fig:different_ginis} shows, the development for the three defined measures of income is quit parallel except for the 1980s. In this time periode the Gini-coefficent for net income veers. This has probably to do with a change in regulations of deductions and shows that interpretation over time has to be very carefull, becauses changes in taxation or regulation systems can affect the outcome. In generall inequality assesst with taxable income is higher than inequality assesst with net income or taxable income after federal taxes. While it is not surprising that federal taxes reduces inequality slightly because of the progressivity of the taxes and it is somehow astonishing that adding deductions (from net income to taxable income) increases inequality.

@@

The main issue concerning the unit of analysis is not easy solved with tax-data, because the concepts of tax-units and households are not perfectly congruent (we will take up this issue, when comparing survey data and the FTA tax data later on.) But we can examine, how the assessment of inequality is affected by the implementation of equivalence scale. We do this by looking and Gini-time series for net income with and without implementation of a equivalence scale. The scale is constructed by using information out of tax data. The income of single household/tax unit are divided by 1 (no change), for married the equivalence-factor is 1.5. For every child or other by the tax-unit supported person a value of 0.3 is added. This measures are provided directly by the FTA and are not calculated by us. Because excluding the group of not-taxed leads to a longer time-series we provide four time-series in total (two possibilities to compare the effect of a equivalence scale)

@@

Here we examine how interpretation can change, when we expand the analysis by using relative distribution methods and not only look at Gini-Coefficients. We therefore use the published percentiles about the distribution of taxable income  on the FTA webpage . We prefer these measures over the calculates measures out of the published income brackets statistics, because they represent the distribution at both tails more accurate since they are based directly on the information about every single tax units\footnote{When calculating percentiles out of the income bracket statistic we loose relevant information at the edges. First, we don't have information about taxable income of tax-units falling below the income threshold for federal taxation. We only know how many persons fall in this category. However, the percentiles reported on the FTA webpage are based on the true taxable income (also for units below the threshold), which allows a more precise estimation of the lower percentiles. Secondly, it is especially hard to estimate the highest top income percentiles out of the aggregated tax statistics, leaving us with information only until the 95\%-percentiles, while the reported percentiles reach the 99.99\%-percentiles}. We use the reported Measures at the cost of time. The longest time-period we can compare out of these data reach from 2003 to 2010. Gini changed during this time from 0.47 to 0.50, which equals a moderate increase of inequality. The in-depth distributional analysis allows us to see, how this change translated into different shapes of the distributions. \\

@@

# Overall rel dist und median shif effect werden deaktiviert
# g10 <- reldist(y=bd.2010.ind$inc, yo=bd.2003.ind$inc,
#               smooth=0.4, ci=FALSE,
#              ylim=c(0.5,2),
#               yolabs=seq(-1,3,by=0.5),
#               bar=TRUE, quiet=FALSE,
#               xlab="Proportion of 2003 Distribution")
#title(main=paste("Overall realtive density"))
#abline(h=1,lty=2)
#g1A <- reldist(y=bd.2010.ind$inc, yo=bd.2003.ind$inc,
#              show="effect",
#               bar=TRUE, quiet=FALSE,
#               ylim=c(0.5,2), ylab="",
#               smooth=0.4, ci=FALSE,
#               yolabs=seq(-1,3,by=0.5),
#               xlab="Proportion of 2003 Distribution")
#title(main=paste("Effect of Median shift"))
#abline(h=1,lty=2)
gA0 <- reldist(y=bd.2010.ind, yo=bd.2003.ind,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2), ylab="",
               discrete=FALSE,
               xlab="Proportion of 2003 Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))
@

@@

@@

# Relative Distribution - Effect of different shapes
c<-reldist(y=pooled_i, yo=normal_i,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.0), ylab="",
               xlab="Proportion of Normal-Case-Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))
@

@@

@@

gA0 <- reldist(y=habe10.married$inc_habe, yo=bd.2010.ind,
               ywgt=habe10.married$gew,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.0), ylab="",
               xlab="Proportion of FTA Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))
@

@@

gA0 <- reldist(y=bd.2010.ind, yo=habe10.unmarried$inc_habe,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.5), ylab="",
               xlab="Proportion of HBS Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))
@

>>>>

In this section we will present results following the strategy introducded in the last section. As describded this includes changeing possible concepts within one field (income definition, statistical unit, measurement and population coverage) while holding other conceptional differences constant. To get the best possible comparison we rely on different data sources (FTA-tax tax data and HBS-survey data) and different time periodes. The presentation of results therefore includes a description of the used data.

@@

We start with the effect of income definition on the assessment of income inequality. As described in the previous section income definitions are either closer to the market outcome or the income, which shapes the possibility to consume (disposable income). With FTA-Tax data in-between income measures can be defined. This includes

@@

As \ref{fig:different_ginis} shows, the development for the three defined measures of income is quite parallel except for the 1980s. In this time periode the Gini-coefficent for net income veers. This has probably to do with a change in regulations of deductions and shows that interpretation over time has to be very careful, becauses changes in taxation or regulation systems can affect the outcome. In general inequality assessed with taxable income is higher than inequality assessed with net income or taxable income after federal taxes. While it is not surprising that federal taxes reduces inequality slightly because of the progressivity of the taxes it is somehow astonishing that adding deductions (from net income to taxable income) increases inequality. This implies deductions to be rather flat not progressive.

@@

The main issue concerning the unit of analysis is not easy to solve with tax-data, because the concepts of tax-units and households are not perfectly congruent (we will take up this issue, when comparing survey data and the FTA tax data later on.) However we can examine, how the assessment of inequality is affected by the implementation of equivalence scale. We do this by looking at Gini-time series for net income with and without implementation of a equivalence scale. The scale is constructed by using information out of tax data. The incomes of single households are divided by 1 (no change), for married tax units the equivalence-factor is 1.5. For every child and person supported by the tax-unit a value of 0.3 is added to the denominator. These measures are provided directly by the FTA and are not calculated by us. Because excluding the group of not-taxed leads to a longer time-series we provide four time-series in total (two possibilities to compare the effect of a equivalence scale).

@@

Here we examine how interpretation can change, when we expand the analysis by using relative distribution methods and not only look at Gini-Coefficients. We therefore use the published percentiles about the distribution of taxable income  on the FTA webpage . We prefer these measures over the calculates measures out of the published income brackets statistics, because they represent the distribution at both tails more accurate since they are based directly on the information about every single tax units\footnote{When calculating percentiles out of the income bracket statistic we loose relevant information at the edges. First, we don't have information about taxable income of tax-units falling below the income threshold for federal taxation. We only know how many persons fall in this category. However, the percentiles reported on the FTA webpage are based on the true taxable income (also for units below the threshold), which allows a more precise estimation of the lower percentiles. Secondly, it is especially hard to estimate the highest top income percentiles out of the aggregated tax statistics, leaving us with information only until the 95\%-percentiles, while the reported percentiles reach the 99.99\%-percentiles}. We use the reported Measures at the cost of time. The longest time-period we can compare out of these data reaches from 2003 to 2010. Gini changed during this time from 0.47 to 0.50, which equals a moderate increase of inequality. The in-depth distributional analysis allows us to see, how this change translated into different shapes of the distributions. \\

@@

# Overall rel dist und median shif effect werden deaktiviert
# g10 <- reldist(y=bd.2010.ind$inc, yo=bd.2003.ind$inc,
#               smooth=0.4, ci=FALSE,
#              ylim=c(0.5,2),
#               yolabs=seq(-1,3,by=0.5),
#               bar=TRUE, quiet=FALSE,
#               xlab="Proportion of 2003 Distribution")
#title(main=paste("Overall realtive density"))
#abline(h=1,lty=2)
#g1A <- reldist(y=bd.2010.ind$inc, yo=bd.2003.ind$inc,
#              show="effect",
#               bar=TRUE, quiet=FALSE,
#               ylim=c(0.5,2), ylab="",
#               smooth=0.4, ci=FALSE,
#               yolabs=seq(-1,3,by=0.5),
#               xlab="Proportion of 2003 Distribution")
#title(main=paste("Effect of Median shift"))
#abline(h=1,lty=2)
gA0 <- reldist(y=bd.2010.ind, yo=bd.2003.ind,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2), ylab="",
               discrete=FALSE,
               xlab="Proportion of 2003 Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))

# Export data to use with Stata
bd2010_2003 <- rbind(data.frame(cbind("bd" = bd.2010.ind,"group"=2010)),data.frame("bd" = bd.2003.ind,"group"=2003))
write.dta(bd2010_2003, file="../../data/stata_data/bd2010_2003.dta")
@

@@

# Export data to use with Stata
ns1993_94 <- rbind(data.frame(cbind("income" = pooled_i,"group"=0)),data.frame("income" = normal_i,"group"=1))
ns1993_94$group <- factor(ns1993_94$group,labels=c("pooled","normal"))
write.dta(ns1993_94, file="../../data/stata_data/ns1993_94.dta")

@@

# Relative Distribution - Effect of different shapes
c<-reldist(y=pooled_i, yo=normal_i,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.0), ylab="",
               xlab="Proportion of Normal-Case-Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))

# Export data to use with Stata
ns2010 <- rbind(data.frame(cbind("income" = pooled_i,"group"=0)),data.frame("income" = normal_i,"group"=1))
ns2010$group <- factor(ns2010$group,labels=c("pooled","normal"))
write.dta(ns2010, file="../../data/stata_data/ns2010.dta")

@@

# Export data to use with Stata
habe_bd_2010 <- rbind(data.frame(cbind("income" = habe10$psteink,"group"=0,"weight"=habe10$gew)),data.frame("income" = bd.2010.ind,"group"=1,"weight"=1))
habe_bd_2010$group <- factor(habe_bd_2010$group,labels=c("habe","bd"))
write.dta(habe_bd_2010, file="../../data/stata_data/habe_bd_2010.dta")

@@

gA0 <- reldist(y=habe10.married$inc_habe, yo=bd.2010.ind,
               ywgt=habe10.married$gew,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.0), ylab="",
               xlab="Proportion of FTA Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))

# Export data to use with Stata
habe_bd_2010_married <- rbind(data.frame(cbind("income" = habe10.married$inc_habe,"group"=0,"weight"=habe10.married$gew)),data.frame("income" = bd.2010.ind,"group"=1,"weight"=1))
habe_bd_2010_married$group <- factor(habe_bd_2010_married$group,labels=c("habe","bd"))
write.dta(habe_bd_2010_married, file="../../data/stata_data/habe_bd_2010_married.dta")

@@

gA0 <- reldist(y=bd.2010.ind, yo=habe10.unmarried$inc_habe,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.5), ylab="",
               xlab="Proportion of HBS Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))

# Export data to use with Stata
habe_bd_2010_unmarried <- rbind(data.frame(cbind("income" = habe10.unmarried$inc_habe,"group"=0)),data.frame("income" = bd.2010.ind,"group"=1))
habe_bd_2010_unmarried$group <- factor(habe_bd_2010_unmarried$group,labels=c("habe","bd"))
write.dta(habe_bd_2010_unmarried, file="../../data/stata_data/habe_bd_2010_unmarried.dta")
